---
title: "DATA 621 Business Analytics & Data Mining" 
subtitle: "Homework #2 Classification Metrics"
author: "Kyle Gilde"
date: "3/11/2018"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    toc_depth: 3
---


```{r knitr_options, echo=FALSE}
knitr::opts_chunk$set(
                      error = FALSE
                      ,message = FALSE
                      #,tidy = TRUE
                      ,cache = TRUE
                      )
```


```{r packages, echo=F, collapse=T} 
installed_and_loaded <- function(pkg){
  # Load packages. Install them if needed.
  # CODE SOURCE: https://gist.github.com/stevenworthington/3178163
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE, quietly = TRUE, warn.conflicts = FALSE)
}

# required packages
packages <- c("prettydoc","tidyverse") 

#excute function and display the loaded packages
data.frame(installed_and_loaded(packages))
```
# Overview

In this homework assignment, you will work through various classification metrics. You will be asked to create functions in R to carry out the various calculations. You will also investigate some functions in packages that will let you obtain the equivalent results. Finally, you will create graphical output that also can be used to evaluate the output of classification models, such as binary logistic regression.

# Deliverables

Upon following the instructions below, use your created R functions and the other packages to generate the classification metrics for the provided data set. A write-up of your solutions submitted in PDF format.

# Instructions

**Complete each of the following steps as instructed:**

## 1. Download the classification output data set (attached in Blackboard to the assignment).

```{r getdata}
class_df <- read.csv("https://raw.githubusercontent.com/kylegilde/D621-Data-Mining/master/HW2%20Classification%20Metrics/classification-output-data.csv")

```

## 2. The data set has three key columns we will use:

+ class: the actual class for the observation 

+ scored.class: the predicted class for the observation (based on a threshold of 0.5) 

+ scored.probability: the predicted probability of success for the observation

Use the table() function to get the raw confusion matrix for this scored dataset. Make sure you
understand the output. In particular, do the rows represent the actual or predicted class? The
columns?

**In the confusion matrix below, the rows represent the actual classes, and the columns represent the predicted classes.**

```{r table}
class_vars <- subset(class_df, select = c(class, scored.class))

table(class_vars)

```


## 3. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the accuracy of the predictions.

```{r Q3}
accuracy <- function(df){
  # Takes a 2-column df where the 1st column is the actual class
  # The 2nd column is the predicted class
  # Calculates the number of true positives & negatives
  # Returns the accuracy rate, the proportion of true predictions
  TP <- sum(df[, 1] == 1 & df[, 2] == 1)
  TN <- sum(df[, 1] == 0 & df[, 2] == 0)
  (TP + TN)/nrow(df)
}


(accuracy_value <- accuracy(class_vars))
```



## 4. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the classification error rate of the predictions.

```{r Q4}
classification_error_rate <- function(df){
  # Takes a 2-column df where the 1st column is the actual class
  # The 2nd column is the predicted class
  # Calculates the number of false positives & negatives
  # Returns the classification error rate, the proportion of false predictions
  FP <- sum(df[, 1] == 0 & df[, 2] == 1)
  FN <- sum(df[, 1] == 1 & df[, 2] == 0)
  (FP + FN)/nrow(df)
}

(cer <- classification_error_rate(class_vars))
```

Verify that you get an accuracy and an error rate that sums to one. 

**They do sum to 1.**

```{r}
cer + accuracy_value == 1
```

5. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the precision of the predictions.


```{r Q5}
precision <- function(df){
  # Takes a 2-column df where the 1st column is the actual class
  # The 2nd column is the predicted class
  # Calculates the number of true & false positives
  # Returns the precision rate, the proportion of the positives that were true
  TP <- sum(df[, 1] == 1 & df[, 2] == 1)
  FP <- sum(df[, 1] == 0 & df[, 2] == 1)
  TP/(TP + FP)
}

(precision_value <- precision(class_vars))
```


6. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the sensitivity of the predictions. Sensitivity is also known as recall.

```{r Q6}
sensitivity <- function(df){
  # Takes a 2-column df where the 1st column is the actual class
  # The 2nd column is the predicted class
  # Calculates the number of true positives & false negatives
  # Returns the sensitivity rate
  # AKA the true positive rate & the probability of detection
  TP <- sum(df[, 1] == 1 & df[, 2] == 1)
  FN <- sum(df[, 1] == 1 & df[, 2] == 0)
  TP/(TP + FN)
}

(sensitivity_value <- sensitivity(class_vars))
```


7. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the specificity of the predictions.

```{r Q7}
specificity <- function(df){
  # Takes a 2-column df where the 1st column is the actual class
  # The 2nd column is the predicted class
  # Calculates the number of true positives & false negatives
  # Returns the specificity rate, 
  # AKA the true negative rate & the proportion of correctly identified negatives
  TN <- sum(df[, 1] == 0 & df[, 2] == 0)
  FP <- sum(df[, 1] == 0 & df[, 2] == 1)
  TN/(TN + FP)
}

(specificity_value <- specificity(class_vars))
```

8. Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the F1 score of the predictions.

```{r Q8}

F1_score <- function(df){
  # Takes a 2-column df where the 1st column is the actual class
  # The 2nd column is the predicted class
  # Calculates the precision & sensitivity
  # Returns the F1 score
  precision_value <- precision(class_vars)
  sensitivity_value <- sensitivity(class_vars)
  (2 * precision_value * sensitivity_value)/(precision_value + sensitivity_value)
}

(F1_score_value <- F1_score(class_vars))

```


9. Before we move on, let's consider a question that was asked: What are the bounds on the F1 score? Show that the F1 score will always be between 0 and 1.


```{r Q9}


arange <- seq(0, 1, by = .05)

possibilities <- expand.grid(arange, arange)

possible_inputs <- merge(possibilities, arange, all = TRUE)
names(possible_inputs)[3] <- "Var3"

F1_score_v2 <- function(values){
  # Takes a 3-value vector of true positive, false positive & false negative values
  # Calculates the precision & sensitivity
  # Returns the F1 score
  iterations <- nrow(values)
  possible_F1_scores <- rep(NA, iterations)

  for (i in 1:iterations){
    TP <- values[i, 1]
    FP <- values[i, 2]
    FN <- values[i, 3]
  
    precision_value <- TP/(TP + FP)
    sensitivity_value <- TP/(TP + FN)
    possible_F1_scores[i] <- (2 * precision_value * sensitivity_value)/(precision_value + sensitivity_value)
  }
    return(possible_F1_scores)
}


F1 <- F1_score_v2(possible_inputs)


plot(F1)

min(F1)
max(F1)

```



10. Write a function that generates an ROC curve from a data set with a true classification
column (class in our example) and a probability column (scored.probability in our example).
Your function should return a list that includes the plot of the ROC curve and a vector that
contains the calculated area under the curve (AUC). Note that I recommend using a sequence of
thresholds ranging from 0 to 1 at 0.01 intervals.
11. Use your created R functions and the provided classification output data set to produce all of
the classification metrics discussed above.
12. Investigate the caret package. In particular, consider the functions confusionMatrix,
sensitivity, and specificity. Apply the functions to the data set. How do the results compare with
your own functions?
13. Investigate the pROC package. Use it to generate an ROC curve for the data set. How do the
results compare with your own functions?

